---
title: "Understanding panelmap"
output: rmarkdown::github_document
---

<!-- generated by README.Rmd -->
<!-- badges: start -->
  [![Build Status](https://travis-ci.org/arorarshi/panelmap.svg?branch=master)](https://travis-ci.org/arorarshi/panelmap)
  [![Codecov test coverage](https://codecov.io/gh/arorarshi/panelmap/branch/master/graph/badge.svg)](https://codecov.io/gh/arorarshi/panelmap?branch=master)
<!-- badges: end -->

```{r, echo = FALSE}
library(knitr)
library(panelmap)
knitr::opts_chunk$set(collapse=TRUE, fig.retina=2, fig.path = "README_figures/README-")
```

## Installation

```{r, eval=FALSE}
library(devtools)
install_github("arorarshi/panelmap")

```

## Requirements 
_panelmap_ requires the following packages - `coin, plyr and circlize`

## Introduction

_panelmap_ creates panels for summarizing binary data, categorical data (more than 2 categories) and continuous data for known groups. Groups can arise from classification algorithms such as kmeans, or other class labels as provided by the user. _panelmap_ is a visual aid to understand associations and trends in data instead from tedious tables and trying to infer multiple plots. 

Each row of  _panelmap_ is called a _panelet_ (singular). Many panelets assemble together to make a _panelmap_ . We first begin by making a panelet for the solution. See `panelet_group`. Rest of the variables are plotted according to the nature of the data - discrete (binary/categorical) - See `panelet_category` , or continuous - See `panelet_continuous` . The function, `makepanel` , is a wrapper function of the above mentioned functions. 

`makepanel` is a wrapper function that outputs a legend, an association table and the panelmap, all with one command. You need to provide the meta information of your data set like, colors, data type, labels color etc. 

We will use `mtcars` data set to illustrate the usage of panelmap. 

See below an example of `panelmap`. 

## Example - panelmap 

We will illustrate _panlemap_ by sorting mtcars data according to variable `cyl` and plotting `mpg, vs, am` and `gear` to understand the relationship between them. 
```{r,fig.width=8, fig.height=4}

#sort everything by cyl
mtcars.sort = mtcars[order(mtcars$cyl),]
#otehr features to plot
mat = mtcars.sort[,c("mpg","vs","am","gear")]

#color assigned to each feature
mat.col=list(am=c("white","black"), vs=c("white","black"), gear=c("yellow","orange","brown"), mpg=c("white", "darkred"))

#type of each feature. type=1 for discrete and type=2 for continuous. 
mat.type= c(2,1,1,1)
tab=makepanel(soln=mtcars.sort$cyl, soln.name="cyl",soln.col=c("red","blue","green"), mat=mat, mat.col=mat.col, mat.type=mat.type, border=TRUE, legend=TRUE, get.pval=TRUE)
```

One can see, how all the variables align with vehicles with 4,6 and 8 cylinder engines, all at a single glance!

`vs` - Binary variable signaling the engine cylinder configuration a V-shape (vs=0) or Straight Line (vs=1). V

`am` - A binary variable signaling whether vehicle has automatic (am=0) or manual (am=1) transmission configuration.

```{r}
kable(tab, caption="Association table corresponding to panelmap")
```

## Example - circomap 

_circomap_ is a visualizing tool to plot and analyze multiple panelmaps, inspired by the circlize^1^ package. 

To illustrate the functionality of _circomap_ we will draw the above panelmap three times in a circular layout. 

```{r,fig.width=8, fig.height=8, message=F}
#dataset 1
dat1 = mtcars[order(mtcars$cyl),]
#dataset 2
dat2 = mtcars[order(mtcars$cyl),]
#dataset 3
dat3 = mtcars[order(mtcars$cyl),]

#put all the datasets in a list 
datasets<-list()
datasets[[1]]=dat1; datasets[[2]]=dat2; datasets[[3]]=dat3
names(datasets) = c("dat1", "dat2", "dat3")
#names of datasets will be used to labek the datasets

#group variable to plot same as panelet_group
gtoplot<-c("cyl")
gcol = c("red","blue","green")
gheight = 0.10 #track height of the group variable 

#features to plot - same as in panelmap example 
ftoplot<-c("mpg", "vs", "am", "gear")
ftype= c(2,1,1,1)
fcol=list(am=c("white","black"), vs=c("white","black"), gear=c("yellow","orange","brown"), mpg=c("white", "darkred"))

fheight<-list(); fheight[1:length(ftoplot)] = 0.08
#Voila!! 
circomap(datasets, gtoplot, gcol, gheight, ftoplot, ftype, fcol, fheight)


```



## Reference
1. Gu, Z. circlize implements and enhances circular visualization in R. Bioinformatics 2014.
