---
title: "Tutorial: Function and usage panelmap"
author: "Arshi Arora"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: Function and usage panelmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>"
)
```


```{r message=FALSE, include=FALSE}

library(panelmap)
library(beeswarm)
library(ggsci)
library(knitr)
library(kableExtra)
library(corrplot)
library(RColorBrewer)
knitr::opts_chunk$set(echo=TRUE, results='asis', message=FALSE, comment=NA, fig.align="center")

```

## Motivation 

Let's see a demonstration examples from the $mtcars$ datasets. 

Performing data exploration by groups or between two variables can be done by boxplot or correlation plots. Let's see a boxplot between variables `mpg` and `cyl`, and a correlation plot between all the variables of `mtcars` dataset. 

We see how `mpg` decreases as number of `cyl` increases. The same relationship can also be seen in the correlation plot at column 8, row 4. 

This can get tedious, and a more visually appealing approach is required to understand the relationship between the variables at a glance. 

```{r bxpl, fig.width=4, fig.height=4}

print(kable(head(mtcars), "html", caption="mtcars dataset") %>%
        kable_styling(bootstrap_options = "striped", full_width = F,position="left"))


#mtcars: mpg vs cyl
boxplot(mpg ~ cyl, data =mtcars, outline = FALSE, main = 'mtcars: cyl vs mpg', varwidth=T, cex.axis=1.5, cex.main=2, frame.plot=F,  xlab="cyl", ylab="mpg", cex.lab=1.5)
beeswarm(mpg ~ cyl, data = mtcars, col = "cadetblue", pch = 16,cex=1.2, add = TRUE)
  
#or a correlation plot 
M <-cor(mtcars)
corrplot(M, type="upper", order="hclust",col=brewer.pal(n=8, name="RdYlBu"))
soln.col = pal_nejm("default")(7)

```

## Example 1 

We will illustrate _panlemap_ by sorting mtcars data according to variable `cyl` and plotting `mpg, vs, am` and `gear` to understand the relationship between them. 

```{r, fig.height=4, fig.width=8}

#mtcars 
#sort everything by cyl
mtcars.sort = mtcars[order(mtcars$cyl),]

#other features to plot
mat = mtcars.sort[,c("mpg","vs","am","gear")]

#color assigned to each feature
mat.col=list(am=c("white","black"), vs=c("white","black"), gear=c("yellow","orange","brown"), mpg=c("white", "darkred"))

#type of each feature. type=1 for discrete and type=2 for continuous. 
mat.type= c(2,1,1,1)
tab=makepanel(gr=mtcars.sort$cyl, gr.name="cyl",gr.col=soln.col[1:3], mat=mat, mat.col=mat.col, mat.type=mat.type, border=TRUE, legend=TRUE, get.pval=TRUE, cex=1.5)
print(kable(tab, "html", caption="panelmap table") %>%
        kable_styling(bootstrap_options = "striped", full_width = F,position="left"))
```

Continuous variable `mpg` is represented as a spectrum of colors between white(low) and darkred(high). See how `mpg` decreases from `cyl` 4-6. 

Other observations that we can make are - 

* cars with `cyl` 3 and 4 are mostly S shape `vs=1`, mostly manual `am=1` and have about 4-5 gears. 

* cars with 5 cylinders `cyl` are mostly V shape `vs=0`, mostly automatic `am=0` and have 3 gears. 

All this at one look!

To compare it with the full correlation matrix, we plot all the features. 

__Taking more features__ 
```{r, fig.height=10, fig.width=10}
#taking more features 
mat.col=list(am=c("white","black"), vs=c("white","black"), gear=c("yellow","orange","brown"), mpg=c("white", "darkred"), disp=c("white", "black"), hp=c("white","black"), drat = c("white", "black"), wt=c("white", "darkblue"), carb=brewer.pal(n=6, name="BuGn") )

mat = mtcars.sort[,names(mat.col)]

#type of each feature. type=1 for discrete and type=2 for continuous. 
mat.type=c(1,1,1,2,2,2,2,2,1)
tab=makepanel(gr=mtcars.sort$cyl, gr.name="cyl",gr.col = soln.col[1:3], mat=mat, mat.col=mat.col, mat.type=mat.type, border=TRUE, legend=TRUE, get.pval=TRUE,lncol=8,lwd=2, cex=1.5, lcex=1.5)

```


## Functions and Usage 

There are three core functions of `panelmap`.

### panelet_group

Given a solution, make a group panelet. Let `cyl` variable with values 4,6 and 8 be our group variable from package `mtcars`. Let's assign the colors `red`, `blue` and `green` to its values as follows :- 

```{r, echo=T, fig.width=8, fig.height=2}
#adjust margins and number of panelet values 
par(mfrow=c(3,1),mar=c(0,8,0,14))

#sort the entire data frame according to the group solution you are interested in 
mtcars.sort = mtcars[order(mtcars$cyl),]

#plot group
tt=panelet_group(gr=mtcars.sort$cyl, gr.col=c("red","blue","green"),gr.name="cyl",cex=1, border=FALSE, legend=FALSE)

```


`cex` controls the magnification of text labels on the left-hand side and is passed as part of `(...)`. Other relevant parameters that can be passed are `col` - to change color of text labels. `lty` and `lwd` to change border line type and magnification, as in baseR. `border` and `legend` take logical values and their default value is `FALSE`. 

Once we have a group panelet, we can start plotting other variables of interest. 

### panelet_category

This function plots discrete data type. Usage - `function(pp,pp.col,soln, var.n="var.n"`

`NA.flag=FALSE, NA.col="grey", get.pval=FALSE, border=FALSE, border.col="black", legend=FALSE,...)` 

These are the default values to the function. There is an option to assign a color for `NAs` in the data by using `NA.flag` and `NA.col`. User can also control the color of border colors by `border.col` 

```{r, echo=T,fig.width=8, fig.height=4, message=F}
#adjust margins and number of panelet values 
par(mfrow=c(6,1),mar=c(0,8,0,14))

#sort the entire data frame according to the group solution you are interested in 
mtcars.sort = mtcars[order(mtcars$cyl),]

#plot group
tt=panelet_group(gr=mtcars.sort$cyl, gr.col=c("red","blue","green"),gr.name="cyl",cex=1, border=FALSE, legend=FALSE)

#plot variable am
tt=panelet_category(pp=mtcars.sort$am, pp.col=c("white", "black"), gr=mtcars.sort$cyl, var.n="am", get.pval=TRUE)

#plot variable vs
tt=panelet_category(mtcars.sort$vs, c("white", "black"), mtcars.sort$cyl, "vs", get.pval=TRUE)

```

`panelet_category` also returns the summary of the variable with respect to the group variable along with an association test. 


### panelet_continuous 

This function provides a gradient of the values present in the continuous variable. Note that this function plots a gradient and assigns colors relative to the range and magnitude of values in the variable. If the values are a mix of negative and positive values the gradient might look biased and lack in contrast. We will try to address this in a future functionality. 

```{r, echo=T,fig.width=8, fig.height=4}
#adjust margins and number of panelet values 
par(mfrow=c(6,1),mar=c(0,8,0,14))

#sort the entire data frame according to the group solution you are interested in 
mtcars.sort = mtcars[order(mtcars$cyl),]

#plot group
tt=panelet_group(gr=mtcars.sort$cyl, gr.col=c("red","blue","green"),gr.name="cyl",cex=1, border=TRUE, legend=FALSE)

#plot variable mpg
tt =panelet_continuous(mtcars.sort$mpg, pp.col=c("white","darkred"), gr=mtcars.sort$cyl,var.n="mpg",NA.flag=FALSE, get.pval=TRUE, border=TRUE)
                       
```

### makepanel 

`makepanel` is a wrapper function that outputs a legend, an association table and the panelmap, all with one command. You need to provide the meta information of your data set like, colors, data type, labels color etc. See [Example 1]

## Other features

### custom visualizations

```{r, fig.height=2, fig.width=10}
#adjust margins and number of panelet values 
par(mfrow=c(4,1),mar=c(0,8,0,14))
#sort the entire data frame according to the group solution you are interested in

#use sorted mtcars 
#plot group 
tt=panelet_group(gr=mtcars.sort$cyl, gr.col=c("red","blue","orange"),gr.name="cyl", cex=2, col="red", border=TRUE)

#plot variable mpg
tt =panelet_continuous(mtcars.sort$mpg, pp.col=c("white","darkred"), gr=mtcars.sort$cyl,var.n="mpg", cex=1.5, col="black", border=TRUE,border.col="green", lwd=3, lty=2)

tt=panelet_category(pp=mtcars.sort$am, pp.col=c("white", "black"), gr=mtcars.sort$cyl, var.n="am", cex=2, col="blue", border=TRUE,border.col="red", lwd=4, lty=3, font=4)

```


### handle NAs or missing data

```{r,fig.height=5, fig.width=10 }
am.na = mtcars.sort$am
am.na[c(1,10,20,30)] = NA
mtcars.sort$am.na = am.na
mat = mtcars.sort[,c("vs","am","am.na","gear")]
mat.col=list(am=c("white","black"), am.na=c("white", "black"), vs=c("white","black"), gear=c("yellow","orange","brown"))

mat.type= c(1,1,1,1)
tab=makepanel(gr=mtcars.sort$cyl, gr.name="cyl",gr.col=soln.col[1:3], mat=mat, mat.col=mat.col, mat.type=mat.type, border=TRUE, legend=TRUE, get.pval=TRUE, cex=1.7, NA.flag=TRUE, NA.col="grey",lncol = 7, lcex=1.3)

print(kable(tab, "html", caption="panelmap NA examaple") %>%
        kable_styling(bootstrap_options = "striped", full_width = F,position="left"))
```


See how there are grey bars in panelet row __am.na__ corresponding to NA at values 1,10,20 and 30. 

